Models of VOR learning in MHC knockout mice

### 2013-02-15 Notes/VORmodel.tex
renamed linear and step multistate models

We will see if we can capture the essential features of VOR gain-increase learning in wild type and MHC \KO\ mutants and wild-type, with and without gain-decrease pre-training.
We will try to do this using models of complex synapses, such as the cascade model \cite{Fusi2005cascade} or the multistate model \cite{amit1994learning,Fusi2007multistate}.
We will also construct a new model in \autoref{sec:pooledmodel} that incorporates the idea that biasing the synapses in one direction could deplete some resource, thereby impairing further plasticity in the same direction in nearby synapses.

@@

On the other hand, the cascade, multistate and pooled resource models all have parameter ranges where the mutant has impaired learning without pre-training (see \autoref{fig:cascade_short}, \autoref{fig:cascade_long}, \autoref{fig:multistate_weak}, \autoref{fig:multistate_strong} and \autoref{fig:pooled_scarce}).
We have found parameter ranges where the reverse happens for the multistate (see discussion below \eqref{eq:multiKOnopre}) and the pooled resource model (see \autoref{fig:pooled_plenty}).
These rely on properties of the mutation, so it would be difficult to explore with further experiments.

All of these models show pre-training enhancing learning for the mutant, as expected.
It is harder to find regimes where pre-training impairs learning for the wild-type.
In the cascade model, this required a much longer duration for pe-training (see \autoref{fig:cascade_long}).
In the multistate model, this required stronger training, which resulted in gain-decrease learning being faster for the mutant then wild-type (see \autoref{fig:multistate_strong}), unlike what is seen experimentally.
It is possible that these features could be improved by lessening the symmetry between gain-increase and decrease or potentiation and depression, or by allowing the rate of plasticity events to vary.
For the pooled resource model, we haven't found any parameter regime where pre-training impairs learning for the wild-type, and we discuss why it seems unlikely in \autoref{sec:pooled}.

There seems to be two take home messages.
First, impaired learning without pre-training in the mutant requires synaptic complexity, as it is not seen in the simple two-state model.
Second, impaired learning with pre-training in the wild-type requires a form of metaplasticity where repeated potentiation makes subsequent depression less likely.
This is a feature of the cascade model and our version of the multistate model, but not the pooled resource model.

@@

We will look at three different models, the cascade model (see \cite{Fusi2005cascade} and \autoref{fig:models}\ref{fig:cascade_model}), the multistate model\footnote{The version of the multistate model that we'll study is slightly different than the original definition. We will only allow the synaptic weight to take two values, whereas it originally varied linearly across the states. } (see \cite{amit1994learning,Fusi2007multistate} and \autoref{fig:models}\ref{fig:multistate_model}) and the two-state model (which can be thought of as a special case of the previous two, see \autoref{fig:models}\ref{fig:binary_model}).
We will also look at a new, pooled resource model that we will define below in \autoref{sec:pooledmodel}.

@@

\begin{table}
 \begin{center}
  \begin{tabularn}{|l|c|c|c|c|c|c|}
    \cline{1-7}
    % after \\: \hline or \cline{col1-col2} \cline{col3-col4} ...
    Model & \# states & pot,WT dep & \KO\ dep & $\Delta f$ & $r\ttrain$ & $r\tpre$ \\
    \cline{1-7}
    Cascade       & 10 & $x=0.25$ & $x=0.33$ & -0.3  & 50 & 50  &\label{tr:cascade_short} \\
    Cascade       & 10 & $x=0.25$ & $x=0.33$ & -0.3  & 50 & 150 &\label{tr:cascade_long} \\
    Multistate    & 10 & $q=0.6$  & $q=0.8$  & -0.1  & 50 & 50  &\label{tr:multistate_weak} \\
    Multistate    & 10 & $q=0.6$  & $q=0.8$  & -0.17 & 10 & 30  &\label{tr:multistate_med} \\
    Multistate    & 10 & $q=0.6$  & $q=0.8$  & -0.4  & 20 & 20  &\label{tr:multistate_strong} \\
    OG Multistate & 10 & $q=0.6$  & $q=0.8$  & -0.3  & 20 & 20  &\label{tr:multistate_lin} \\
    Two-state     & 2  & $q=0.4$  & $q=0.8$  & -0.1  & 5  & 5   &\label{tr:binary} \\
    Pooled        & 10 & $q\in[0.3,0.4]$  & $q\in[0.6,0.8]$
                                          & -0.1 & 70 & 70 &\label{tr:pooled_plenty}\\
    Pooled        & 10 & $q\in[0.1,0.4]$  & $q\in[0.2,0.8]$
                                          & -0.1 & 70 & 70 &\label{tr:pooled_scarce}\\
    \cline{1-7}
%   \hline
%    % after \\: \hline or \cline{col1-col2} \cline{col3-col4} ...
%    & Model & \# states & pot,WT dep & \KO\ dep & $\Delta f$ & $r\ttrain$ & $r\tpre$ \\
%    \cline{1-7}
%    \label{tr:cascade_short}&
%    Cascade    & 10 & $x=0.25$ & $x=0.33$ & -0.3 & 50 & 50   \\
%    \label{tr:cascade_long}&
%    Cascade    & 10 & $x=0.25$ & $x=0.33$ & -0.3 & 50 & 150  \\
%    \label{tr:multistate_weak}&
%    Multistate & 10 & $q=0.6$  & $q=0.8$  & -0.1 & 50 & 50   \\
%    \label{tr:multistate_strong}&
%    Multistate & 10 & $q=0.6$  & $q=0.8$  & -0.4 & 20 & 20   \\
%    \label{tr:binary}&
%    Two-state  & 2  & $q=0.4$  & $q=0.8$  & -0.1 & 5  & 5    \\
%    \label{tr:pooled_plenty}&
%    Pooled     & 10 & $q\in[0.3,0.4]$  & $q\in[0.6,0.8]$
%                                          & -0.1 & 70 & 70 \\
%    \label{tr:pooled_scarce}&
%    Pooled     & 10 & $q\in[0.1,0.4]$  & $q\in[0.2,0.8]$
%                                          & -0.1 & 70 & 70 \\
%    \cline{1-7}
  \end{tabularn}
 \end{center}
  \caption{Parameters used in simulations.} \label{tab:params}
\end{table}

@@

\subsubsection{Multistate model}\label{sec:multistate}

\resultsfig{Simulation results for the multistate model with weak training}{($\Delta f=-0.1$)}{multistate_weak}

\resultsfig{Simulation results for the multistate model with moderate training}{($\Delta f=-0.17$)}{multistate_med}

\resultsfig{Simulation results for the multistate model with strong training}{($\Delta f=-0.4$)}{multistate_strong}

The results of simulations of the multistate model can be seen in \autoref{fig:multistate_weak}, \autoref{fig:multistate_med} and \autoref{fig:multistate_strong}.
However, we can get some insight into this model analytically.

@@

\subsubsection{OG multistate model}\label{sec:multistate_lin}

>>>>

We will see if we can capture the essential features of VOR gain-increase learning in wild type and MHC \KO\ mutants and wild-type, with and without gain-decrease pre-training.
We will try to do this using models of complex synapses, such as the cascade model \cite{Fusi2005cascade} or the multistate model \cite{amit1994learning,Fusi2007multistate}.\footnote{The original definition of the multistate model had a synaptic weight that varied linearly along the chain. We will refer to this as the linear multistate model. We will also look at the step-multistate model, for which the synaptic weight to take two values.}
We will also construct a new model in \autoref{sec:pooledmodel} that incorporates the idea that biasing the synapses in one direction could deplete some resource, thereby impairing further plasticity in the same direction in nearby synapses.

@@

A similar thing happens for the linear multistate model (see \autoref{fig:multistate_lin}), but the wild-type catches up and overtakes the mutant very quickly.

On the other hand, the cascade, step-multistate and pooled resource models all have parameter ranges where the mutant has impaired learning without pre-training (see \autoref{fig:cascade_short}, \autoref{fig:cascade_long}, \autoref{fig:multistate_weak}, \autoref{fig:multistate_strong} and \autoref{fig:pooled_scarce}).
We have found parameter ranges where the reverse happens for the multistate (see discussion below \eqref{eq:multiKOnopre}) and the pooled resource model (see \autoref{fig:pooled_plenty}).
These rely on properties of the mutation, so it would be difficult to explore with further experiments.

All of these models show pre-training enhancing learning for the mutant, as expected.
It is harder to find regimes where pre-training impairs learning for the wild-type.
In the cascade model, this required a much longer duration for pe-training (see \autoref{fig:cascade_long}).
In the step-multistate model, this required stronger training, which resulted in gain-decrease learning being faster for the mutant then wild-type (see \autoref{fig:multistate_strong}), unlike what is seen experimentally.
It is possible that these features could be improved by lessening the symmetry between gain-increase and decrease or potentiation and depression, or by allowing the rate of plasticity events to vary.
For the linear multistate and pooled resource models, we haven't found any parameter regime where pre-training impairs learning for the wild-type, and we discuss why it seems unlikely in \autoref{sec:pooled}.

There seems to be two take home messages.
First, impaired learning without pre-training in the mutant requires synaptic complexity, as it is not seen in the simple two-state model.
Second, impaired learning with pre-training in the wild-type requires a form of metaplasticity where repeated potentiation makes subsequent depression less likely.
This is a feature of the cascade model and our version of the multistate model, but not the pooled resource model or linear multistate.

@@

We will look at three different models, the cascade model (see \cite{Fusi2005cascade} and \autoref{fig:models}\ref{fig:cascade_model}), the multistate model (see \cite{amit1994learning,Fusi2007multistate} and \autoref{fig:models}\ref{fig:multistate_model}) and the two-state model (which can be thought of as a special case of the previous two, see \autoref{fig:models}\ref{fig:binary_model}).
We will also look at a new, pooled resource model that we will define below in \autoref{sec:pooledmodel}.

@@

\begin{table}
 \begin{center}
  \begin{tabularn}{|l|c|c|c|c|c|c|}
    \cline{1-7}
    % after \\: \hline or \cline{col1-col2} \cline{col3-col4} ...
    Model & \# states & pot,WT dep & \KO\ dep & $\Delta f$ & $r\ttrain$ & $r\tpre$ \\
    \cline{1-7}
    Cascade       & 10 & $x=0.25$ & $x=0.33$ & -0.3  & 50 & 50  &\label{tr:cascade_short} \\
    Cascade       & 10 & $x=0.25$ & $x=0.33$ & -0.3  & 50 & 150 &\label{tr:cascade_long} \\
    Step-multistate    & 10 & $q=0.6$  & $q=0.8$  & -0.1  & 50 & 50  &\label{tr:multistate_weak} \\
    Step-multistate    & 10 & $q=0.6$  & $q=0.8$  & -0.17 & 10 & 30  &\label{tr:multistate_med} \\
    Step-multistate    & 10 & $q=0.6$  & $q=0.8$  & -0.4  & 20 & 20  &\label{tr:multistate_strong} \\
    Linear Multistate & 10 & $q=0.6$  & $q=0.8$  & -0.3  & 20 & 20  &\label{tr:multistate_lin} \\
    Two-state     & 2  & $q=0.4$  & $q=0.8$  & -0.1  & 5  & 5   &\label{tr:binary} \\
    Pooled        & 10 & $q\in[0.3,0.4]$  & $q\in[0.6,0.8]$
                                          & -0.1 & 70 & 70 &\label{tr:pooled_plenty}\\
    Pooled        & 10 & $q\in[0.1,0.4]$  & $q\in[0.2,0.8]$
                                          & -0.1 & 70 & 70 &\label{tr:pooled_scarce}\\
    \cline{1-7}
%   \hline
%    % after \\: \hline or \cline{col1-col2} \cline{col3-col4} ...
%    & Model & \# states & pot,WT dep & \KO\ dep & $\Delta f$ & $r\ttrain$ & $r\tpre$ \\
%    \cline{1-7}
%    \label{tr:cascade_short}&
%    Cascade    & 10 & $x=0.25$ & $x=0.33$ & -0.3 & 50 & 50   \\
%    \label{tr:cascade_long}&
%    Cascade    & 10 & $x=0.25$ & $x=0.33$ & -0.3 & 50 & 150  \\
%    \label{tr:multistate_weak}&
%    Multistate & 10 & $q=0.6$  & $q=0.8$  & -0.1 & 50 & 50   \\
%    \label{tr:multistate_strong}&
%    Multistate & 10 & $q=0.6$  & $q=0.8$  & -0.4 & 20 & 20   \\
%    \label{tr:binary}&
%    Two-state  & 2  & $q=0.4$  & $q=0.8$  & -0.1 & 5  & 5    \\
%    \label{tr:pooled_plenty}&
%    Pooled     & 10 & $q\in[0.3,0.4]$  & $q\in[0.6,0.8]$
%                                          & -0.1 & 70 & 70 \\
%    \label{tr:pooled_scarce}&
%    Pooled     & 10 & $q\in[0.1,0.4]$  & $q\in[0.2,0.8]$
%                                          & -0.1 & 70 & 70 \\
%    \cline{1-7}
  \end{tabularn}
 \end{center}
  \caption{Parameters used in simulations.} \label{tab:params}
\end{table}

@@

\subsubsection{Step-multistate model}\label{sec:multistate}

\resultsfig{Simulation results for the step-multistate model with weak training}{($\Delta f=-0.1$)}{multistate_weak}

\resultsfig{Simulation results for the step-multistate model with moderate training}{($\Delta f=-0.17$)}{multistate_med}

\resultsfig{Simulation results for the step-multistate model with strong training}{($\Delta f=-0.4$)}{multistate_strong}

The results of simulations of the step-multistate model can be seen in \autoref{fig:multistate_weak}, \autoref{fig:multistate_med} and \autoref{fig:multistate_strong}.
However, we can get some insight into this model analytically.

@@

\subsubsection{Linear multistate model}\label{sec:multistate_lin}
